%%% TeX-master: "manuscript"

\section{Figure Legends}

\subsection{Figure 1}

Schematic overview of NAD biosynthesis and consumption. NAD can be synthesized using several routes from altogether three main precursors: tryptophan (Trp), nicotinamide (Nam), and nicotinic acid (NA). Nam and NA are together known as vitamin B3 or niacin. To a lesser extend, nicotinamide ribose (NR) can be used omitting the energetically unfavourable reaction of Nam phosphoribosyltransferase (NamPRT), requiring nicotinamide ribose kinase (NRK) instead. Only 1\% of the tryptophan taken up with our diet is converted into NAD, thus, vitamin B3 and, to a lower extend, NR are essential components of our diet, with Nam being the major NAD precursor in humans\todo{Whole sentence (starting with "Only") not necessarily part of the legend. What do you think, Ines?}. Nam is furthermore the product of NAD-consuming signalling reactions such as sirtuins (NAD-dependent histone deacetylases) or PARPs (poly-ADP-ribosylases).

For the recycling of Nam, two different pathways exist. The pathway found in yeast and many bacteria is using a four-step pathway starting with the deamination of Nam to nicotinic acid (NA) by Nam deamidase (NADA). The other three enzymes comprise the Preiss-Handler pathway that also exists in vertebrates. The recycling pathway found in vertebrates directly converts Nam into the corresponding mononucleotide (NMN) a reaction catalysed by NamPRT and driven by a non-stoichiometric ATP-hydrolysis. A similar reaction catalysed by an evolutionary related enzyme NAPRT, converts NA into the NA mononucleotide (NAMN). NMN and NAMN are converted into dinucleotides by the Nam/NA adenylytransferases (NMNATs). The recycling pathway via NA finally requires an amination step catalysed by NADsynthase (NADS), driven by the conversion of ATP to AMP producing pyrophosphate.


\subsection{Figure 2}

\textbf{Evolutionary distribution of NADA, NNMT and NamPRT and their relation to the number of NAD consumers} {A) Distribution of NADA, NNMT and NamPRT in selected major taxa. NADA is dominant in Bacteria, Fungi, and Plants (Viridiplantae), whereas NamPRT together with NNMT is dominant in Deuterostomia. Numbers at the pie charts show, how many species of the taxon possess the respective enzyme combination indicated by the colour explained in the lower right of the figure. Below the taxon name, the number of species in that taxon is given. B) Common tree of selected taxa within the Metazoa, including 334 species. The pie charts indicate the distribution of species within the respective taxon that have the enzyme combination indicated by the colour, explained in the lower right. The size of the pie charts is proportional to the logarithm of the number of species analysed in the particular taxon. The numbers below the taxon names indicate the average number of NAD-consuming enzyme families found in all sub-taxa. The branch length is arbitrary.


\subsection{Figure 3}

\textbf{NNMT enables high NAD-consumption flux} We used a dynamic model of NAD-biosynthesis and consumption (details see Materials and Methods)  to simulate NAD consumption flux (A) and NAD concentration (B) in the presence of NamPRT and with or without NNMT at different cell division rates.  NMMT enables higher steady state NAD-consumption flux despite reduced NAD-concentrations.


\subsection{Figure 4}
\textbf{Role of NamPRT affinity for Nam} Using the dynamic model of NAD-biosynthesis and consumption we simulated the affect of different Michaelis Menten constants ($K_M$)  of NamPRT for Nam on steady state NAD consumption flux and NAD concentration at different cell devision rates. In the absence of NNMT (A and B) The $K_M$ of NamPRT has little influence on NAD consumption and concentration, but both are changing with cell devision rates. In the presence of NNMT (C and D), decreasing $K_M$ of NamPRT enables incresing NAD consumption flux and NAD concentration. NNMT furthermore makes both relatively independent of cell devision rates. Comparing the situation with and without NNMT (E and F) at different NamPRT $K_M$ reveals that at low $K_M$ and high cell devision rates NNMT no longer enables higher NAD consumption rates compared to NamPRT alone



\subsection{Figure 5}

\textbf{The substrate affinities of human NNMT and NamPRT are optimal}. Both  NAD consumption (A) and NAD concentration (B) are increasing with decreasing $K_M$ of NamPRT but decreeasing $K_M$ of NNMT. The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.

\subsection{Figure 6}

\textbf{The function of the structurally unresolved loop structure of NamPRT} Most deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A). Coloured circles indicate the enzymes present in the species; blue: NamPRT and NNMT; black: NamPRT, NADA, and NNMT; yellow: NamPRT and NADA. The inserted region is not resolved in crystal structures of human NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red  (B) . The visualisation is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} of the sequence of the human NamPRT (P43490) using the model 2H3D as template \cite{Wang2006}.  The localisation of the FLAG-tagged mutant protein is not changed (E) compared to NamPRT-wildtype. Both show a heterogenous nuclear cytosolic localisation in immunofluerescence images in .... cells (C). But, in vitro measurements using recombinant protein show  that a mutant of NamPRT missing this insertion has lower activity (D)  and is not activated by ATP (E). 

\subsection{Figure 7}
\textbf{NNMT provides a competitive advantage and makes NADA obsolete} To simulate competition for common resources we created a two compartment model, where one compartment contained NADA but no NamPRT and one compartment contained NamPRT either with or withour NNMT but no NADA. In the absence of NNMT (A-B) the compartment containing NADA has slightly lower NAD-consumption rates (A) but much higher NAD concentrations (B)  In the presence of NNMT, however, both NAD consumption (C) and NAD concentration (D) are lower in the NADA compartment, but this effect is only observed at low NamPRT $K_M$. 


